---
title: "Rcollectl"
author: 
  - name: Vince Carey
    affiliation:
    - Harvard Medical School
    email: stvjc@channing.harvard.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('Rcollectl')`"
vignette: >
  %\VignetteIndexEntry{Rcollectl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("knitcitations")

## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = "to.doc", citation_format = "text", style = "html")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitcitations = citation("knitcitations")[1],
    knitr = citation("knitr")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    Rcollectl = citation("Rcollectl")[1]
)

write.bibtex(bib, file = "Rcollectl.bib")
```

# Basics

## Install `Rcollectl`

`R` is an open-source statistical environment which can be easily modified to enhance its functionality via packages. `r Biocpkg("Rcollectl")` is a `R` package available via github at `vjcitn/Rcollectl`.

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
  }

BiocManager::install("vjcitn/Rcollectl")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

## Citing `Rcollectl`

We hope that `r Biocpkg("Rcollectl")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("Rcollectl")
```

# Quick start to using to `Rcollectl`

```{r "start", message=FALSE}
library("Rcollectl")
```
Collectl will perform measurements on system resource 
consumption of various types.  We provide a demonstration
output with the package:

```{r lkdemo}
lk = cl_parse(system.file("demotab/demo_1123.tab.gz", package="Rcollectl"))
dim(lk)
attr(lk, "meta")
lk[1:5,1:5]
```

Essentially, `collectl` will generate files like the .tab.gz referenced
above.  `cl_parse` can ingest this information, retaining metadata about
the invocation, into an R object
(in this case, a `data.frame` instance with an extra
attribute).  We will analyze the resource consumption using
the table.  A simple visualization will help fix concepts:

```{r lkviz}
plot_usage(lk)
```

From this display, we can see that about a burst of network activity
around 14:43 is followed by consumption of CPU, memory, and disk resources.
The % CPU active never exceeds 30, memory consumption started relatively
high when sampling began, growing to about 15.5 GB. and 250MB were written
to disk over the entire interval.


# Reproducibility

The `r Biocpkg("Rcollectl")` package `r citep(bib[["Rcollectl"]])` was made possible thanks to:

* R `r citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitcitations")` `r citep(bib[["knitcitations"]])`
* `r CRANpkg("knitr")` `r citep(bib[["knitr"]])`
* `r CRANpkg("rmarkdown")` `r citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Githubpkg("lcolladotor/biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("Rcollectl.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("Rcollectl.Rmd", tangle = TRUE)
```

```{r createVignette2}
## Clean up
file.remove("Rcollectl.bib")
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("knitcitations")` `r citep(bib[["knitcitations"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
bibliography()
```
