---
title: "Rcollectl"
author: 
  - name: Vince Carey
    affiliation:
    - Harvard Medical School
    email: stvjc@channing.harvard.edu
  - name: Yubo Cheng
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: yubo.cheng@roswellpark.org
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('Rcollectl')`"
vignette: >
  %\VignetteIndexEntry{Rcollectl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("knitcitations")

## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = "to.doc", citation_format = "text", style = "html")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitcitations = citation("knitcitations")[1],
    knitr = citation("knitr")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    Rcollectl = citation("Rcollectl")[1]
)

write.bibtex(bib, file = "Rcollectl.bib")
```

# Quick start for `Rcollectl`

```{r "start", message=FALSE}
library("Rcollectl")
```
Collectl will perform measurements on system resource 
consumption of various types.  We provide a demonstration
output with the package:

```{r lkdemo}
lk = cl_parse(system.file("demotab/demo_1123.tab.gz", package="Rcollectl"))
dim(lk)
attr(lk, "meta")
lk[1:5,1:5]
```


```{r lkviz}
plot_usage(lk)
```

From this display, we can see that about a burst of network activity
around 14:43 is followed by consumption of CPU, memory, and disk resources.
The % CPU active never exceeds 30, memory consumption started relatively
high when sampling began, growing to about 15.5 GB. and 250MB were written
to disk over the entire interval.

To generate a display like this, we use commands shown below:

```{r lklk,eval=FALSE}
id = cl_start([target file prefix])
[use R until task to be measured is complete]
cl_stop(id)
usage_df = cl_parse(dir(patt=[target file prefix]))
# analyze or filter the usage_df (for example, to trim away
# time related to task delay or delay of `cl_stop`
plot_usage(usage_df)
```

# Timestamps

Yubo Cheng has added functionality allowing us to annotate
usage plots with labels related to task phases.  Here is
the code from the example showing how to introduce annotations
in the time profile.

```{r lkts, eval=TRUE}
     id <- cl_start() # March 8 blocking as SPB is erroring here
     Sys.sleep(2)
     #code
     cl_timestamp(id, "step1")
     Sys.sleep(2)
     # code
     Sys.sleep(2)
     cl_timestamp(id, "step2")
     Sys.sleep(2)
     # code
     Sys.sleep(2)
     cl_timestamp(id, "step3")
     Sys.sleep(2)
     # code
     cl_stop(id)
     path <- cl_result_path(id)
     
     plot_usage(cl_parse(path)) +
       cl_timestamp_layer(path) +
       cl_timestamp_label(path) +
       ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust=1))
```


# Reproducibility

The `r Biocpkg("Rcollectl")` package `r citep(bib[["Rcollectl"]])` was made possible thanks to:

* R `r citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitcitations")` `r citep(bib[["knitcitations"]])`
* `r CRANpkg("knitr")` `r citep(bib[["knitr"]])`
* `r CRANpkg("rmarkdown")` `r citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Githubpkg("lcolladotor/biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("Rcollectl.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("Rcollectl.Rmd", tangle = TRUE)
```

```{r createVignette2}
## Clean up
file.remove("Rcollectl.bib")
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("knitcitations")` `r citep(bib[["knitcitations"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
bibliography()
```
