---
title: "Single-Cell Analysis Workflow of HCA Loom File"
output:
  html_document: default
params:
  knitr_eval: FALSE
  fileId: "902ac276-d7bf-5ae2-b89d-f80096a0b310"
  sample: "donor_organism.biomaterial_core.biomaterial_id"
vignette: >
  %\VignetteIndexEntry{Single-Cell Analysis Workflow of HCA Loom File}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, eval = params$knitr_eval)
```

This vignette follows single-cell analysis Workflow (workflow chapters of the “Orchestrating Single-Cell Analysis with Bioconductor” book) to analyze loom file from HCA dataset and tracks computing resources usage.

## Set up parameters

File id of loom file to analyze could be obtained using package `hca`, and one example shows below.

```{r, echo = TRUE, message=FALSE}
library(hca)
library(dplyr)
files(filters(
    projectTitle = list(is = "Census of Immune Cells"),
    organPart = list(is = "bone marrow"),
    fileFormat = list(is = "loom"),
    isIntermediate = list(is = FALSE))
)
```

```{r, echo = TRUE, message=FALSE, eval=TRUE}
fileId <- params$fileId
```

Column to analyze needs to be specified from colData of the dataset, potential targets could be donor_organism.biomaterial_core.biomaterial_id, cell_suspension.biomaterial_core.biomaterial_id, etc.

```{r, echo = TRUE, message=FALSE, eval=TRUE}
sample <- params$sample
```

```{r, echo = TRUE, message=FALSE, eval=TRUE}
fileId
sample
```

With two parameters set up, workflow could be started. 

## Start for `Rcollectl` and download loom file

```{r cl_start, echo = TRUE, message=FALSE}
library(Rcollectl)
library(hca)
library(dplyr)

## identify target loom file
myfile <- files(filters(fileId = list(is = fileId)))

## start Rcollectl
id <- cl_start(gsub('.loom', '', myfile$name))

## download file
loom_file <- files_download(myfile)
```

## Step data preparation 

```{r data_preparation, echo = TRUE, message=FALSE}
# record start time of each step
cl_timestamp(id, "data preparation")

## import loom file
library(LoomExperiment)
sce_loom <- LoomExperiment::import(loom_file, type = "SingleCellLoomExperiment")

## Using manifest data to annotate a .loom file
sce_loom <- optimus_loom_annotation(sce_loom)

## pre-process data
library(SingleCellExperiment)
sce_loom$sample <- colData(sce_loom)[, sample]

library(EnsDb.Hsapiens.v86)
rowData(sce_loom)$Chr <- mapIds(EnsDb.Hsapiens.v86, keys=substring(rowData(sce_loom)$ensembl_ids, 1, 15), column="SEQNAME", keytype="GENEID")

library(scater)
rownames(sce_loom) <- uniquifyFeatureNames(substring(rowData(sce_loom)$ensembl_ids, 1, 15), names = rowData(sce_loom)$Gene)

names(assays(sce_loom)) <- "counts"
```

## Step quality control

```{r quality_control, message=FALSE}
cl_timestamp(id, "quality control")

library(BiocParallel)
library(scater)
bpp <- MulticoreParam(parallel::detectCores())
sce_loom <- unfiltered <- addPerCellQC(sce_loom, BPPARAM=bpp, subsets=list(Mito=which(rowData(sce_loom)$Chr=="MT")))
qc <- quickPerCellQC(colData(sce_loom), batch=sce_loom$sample, sub.fields="subsets_Mito_percent")

sce_loom <- sce_loom[,!qc$discard]
unfiltered$discard <- qc$discard

## Distribution of QC metrics in the HCA bone marrow dataset. Each point represents a cell and is colored according to whether it was discarded
gridExtra::grid.arrange(
    plotColData(unfiltered, x="sample", y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, x="sample", y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, x="sample", y="subsets_Mito_percent",
                colour_by="discard") + ggtitle("Mito percent"),
    ncol=2
)

## Percentage of mitochondrial reads in each cell in the HCA bone marrow dataset compared to its total count. Each point represents a cell and is colored according to whether that cell was discarded
plotColData(unfiltered, x="sum", y="subsets_Mito_percent", colour_by="discard") + scale_x_log10()

## Normalization
sce_loom <- logNormCounts(sce_loom, size_factors = sce_loom$sum)
summary(sizeFactors(sce_loom))
```

## Step variance modeling

```{r variance_modeling, message=FALSE}
cl_timestamp(id, "variance modeling")

library(scran)
set.seed(101001)

dec <- modelGeneVarByPoisson(sce_loom, block=sce_loom$sample, BPPARAM=bpp)
top <- getTopHVGs(dec, n=5000)

## Per-gene variance as a function of the mean for the log-expression values in the HCA bone marrow dataset. Each point represents a gene (black) with the mean-variance trend (blue) fitted to the variances
par(mfrow=c(4,2))
par(mar=c(1,1,1,1))
blocked.stats <- dec$per.block
for (i in colnames(blocked.stats)) {
    current <- blocked.stats[[i]]
    plot(current$mean, current$total, main=i, pch=16, cex=0.5,
         xlab="Mean of log-expression", ylab="Variance of log-expression")
    curfit <- metadata(current)
    curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
}
```

## Step data integration

```{r data_integration, message=FALSE}
# This step takes longer time
cl_timestamp(id, "data integration")

library(batchelor)
library(BiocNeighbors)

set.seed(1010001)
merged <- fastMNN(sce_loom, batch = sce_loom$sample, subset.row = top,
                  BSPARAM=BiocSingular::RandomParam(deferred = TRUE), 
                  BNPARAM=AnnoyParam(), BPPARAM=bpp)

reducedDim(sce_loom, 'MNN') <- reducedDim(merged, 'corrected')
metadata(merged)$merge.info$lost.var
```

## Step dimensionality reduction

```{r dimensionality_reduction, message=FALSE}
cl_timestamp(id, "dimensionality reduction")

library(uwot)
set.seed(1010001)
sce_loom <- runUMAP(sce_loom, dimred="MNN",
                    external_neighbors=TRUE, 
                    BNPARAM=AnnoyParam(),
                    BPPARAM=bpp,
                    n_threads=bpnworkers(bpp))
```

## Step clustering

```{r clustering, message=FALSE}
cl_timestamp(id, "clustering")

library(bluster)

set.seed(1010001)
colLabels(sce_loom) <- clusterRows(reducedDim(sce_loom, "MNN"), TwoStepParam(KmeansParam(centers=1000), NNGraphParam(k=5)))

table(colLabels(sce_loom))

tab <- table(Cluster=colLabels(sce_loom), sample=sce_loom$sample)

## Heatmap of log10-number of cells in each cluster (row) from each sample (column)
library(pheatmap)
pheatmap(log10(tab+10), color=viridis::viridis(100))

scrambled <- sample(ncol(sce_loom))

## UMAP plots of the HCA bone marrow dataset after merging. Each point represents a cell and is colored according to the assigned cluster (top) or the sample of origin (bottom)
gridExtra::grid.arrange(
    plotUMAP(sce_loom, colour_by="label", text_by="label"),
    plotUMAP(sce_loom[,scrambled], colour_by="sample")
)
```

## Step differential expression

```{r differential_expression, message=FALSE}
cl_timestamp(id, "differential expression")

markers <- findMarkers(sce_loom, block = sce_loom$sample, direction = 'up', lfc = 1, BPPARAM=bpp)

top.markers <- markers[["4"]]
best <- top.markers[top.markers$Top <= 10,]
lfcs <- getMarkerEffects(best)

## Heatmap of log2-fold changes for the top marker genes (rows) of cluster 4 compared to all other clusters (columns)
pheatmap(lfcs, breaks=seq(-5, 5, length.out=101))
```

## Step cell type classification

```{r cell_type_classification, message=FALSE}
cl_timestamp(id, "cell type classification")

se.aggregated <- sumCountsAcrossCells(sce_loom, id=colLabels(sce_loom), BPPARAM=bpp)

library(celldex)
hpc <- HumanPrimaryCellAtlasData()

library(SingleR)
anno.single <- SingleR(se.aggregated, ref = hpc, labels = hpc$label.main, assay.type.test="sum")
anno.single
```

## Stop `Rcollectl`

```{r cl_stop}
cl_stop(id)
```

## Plot collectl result

```{r plot_all}
path <- cl_result_path(id)
plot_usage(cl_parse(path))
```

## Plot collectl result with time stamp

```{r plot_timestamp}
path <- cl_result_path(id)
plot_usage(cl_parse(path)) +
  cl_timestamp_layer(path) +
  cl_timestamp_label(path) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## `R` session information

```{r session_info, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```
